---
title: "STA 518 Data Transformations & Manipulation with dplyr Activity"
author: "Anto"
date: "2023-10-02"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's load some packages to start

```{r}
library(tidyverse)
library(lubridate)
library(knitr)
library(skimr)
library(scales)
library(dplyr)
library(lubridate)
library(ggplot2)
```

Then, let's import the flights data into R.

```{r}
# Importing flights data
miFlights <- read_csv("C:/Users/antol/Downloads/miFlights2019-2021/miFlights2019-2021.csv")
```

Let's use `glimpse()` and `skim()` to get familiar with the data.

```{r}
glimpse(miFlights)
```

```{r}
skim(miFlights)
```

## Bar chart

*Create a bar chart showing how many flights departed out of each airport (origin) using the `count()` and `geom_col()` functions. Also sort the bars by descending height using the `fct_reorder()` function.*

Let's make a bar chart.

```{r}
# Making table of counts
originCounts <- miFlights %>%
  dplyr::count(origin)

# Making a bar chart
originCounts %>%
  ggplot(aes(x = fct_reorder(origin, n, .desc = TRUE), y = n)) +
  geom_col(fill = "dodgerblue",
           color = "black") +
  scale_y_continuous(labels = scales::label_comma(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Number of flights out of major Michigan airports, 2019-2021",
       x = "Airport",
       y = "Number of flights" ,
       caption = "Data source: anyFlights R package") +
  ggthemes::theme_few()
```

## The `filter()` function

*Selecting all flights on January 1st in the data set, create a new object called janFlights.*

```{r}
# Subsetting to January first flights
janFlights <- miFlights %>%
  dplyr::filter(month == 1, day == 1)
```

*Suppose we want to create a data set called dec25 that contains flight data from December 25th. What code would we need using the filter() function to create dec25?*

```{r}
# Subsetting to January first flights
dec25 <- miFlights %>%
  dplyr::filter(month == 12, day == 25)
```

*Find all flights that departed in November or December, creating an object called novDec.*

```{r}
# Subsetting to flights in November or December
novDec <- miFlights %>%
  dplyr::filter(month == 11 | month == 12)
```

*Find all flights that departed in November or December using the %in% operator, creating an object called novDec.*

```{r}
# Subsetting to flights in November or December
novDec <- miFlights %>%
  dplyr::filter(month %in% c(11:12))
```

*Select all flights except those in the months of November and December using !.*

```{r}
# Subsetting to flights in months except in November or December
notNovDec <- miFlights %>%
  dplyr::filter(!(month %in% 11:12))
```

*Create a new object called miFlightsComplete where all departure times are non-missing, and miFlightsMiss where all departure times are missing*

```{r}
# Only keeping rows with missing values
miFlightsMiss <- miFlights %>%
  dplyr::filter(is.na(dep_time))

# Excluding missing values
miFlightsComplete <- miFlights %>%
  dplyr::filter(!is.na(dep_time))
```

## The `arrange()` function

*Sort miFlights by the day of the flight (smallest to largest), and print the first 4 columns and 5 rows of the resulting data set using the slice_head() function.*

```{r}
# Printing rows of sorted data
miFlights %>%
  arrange(day) %>%
  dplyr::select(1:4) %>%
  slice_head(n = 5)
```

*Sort miFlights by the day of the flight (largest to smallest), and print the first 4 columns and 5 rows of the resulting data set using the slice_head() function.*

```{r}
# Printing rows of sorted data
miFlights %>%
  arrange(desc(day)) %>%
  dplyr::select(1:4) %>%
  slice_head(n = 5)
```

*Sort miFlights by the year, month, and day of the flight.*

```{r}
# Printing rows of sorted data
miFlights <- miFlights %>%
  arrange(year, month, day)
```

1. Sort miFlights to find the 3 most delayed flights (arr_delay), and the 3 that left the earliest relative to their scheduled departure (`dep_delay`).

```{r}
# To find the 3 most delayed flights (arr_delay)
most_delayed <- miFlights %>%
  arrange(desc(arr_delay)) %>%
  slice_head(n = 3)

# To find the 3 flights that left the earliest (dep_delay)
earliest_departures <- miFlights %>%
  arrange(dep_delay) %>%
  slice_head(n = 3)
```

2. Sort miFlights to find the 3 fastest (`highest speed`) flights. 

```{r}
# Calculate speed for each flight (in miles per hour)
miFlights <- miFlights %>%
  mutate(speed = distance / (air_time / 60)) # air_time is in minutes, so dividing by 60 converts it to hours

# Sort the data frame by speed to find the 3 fastest flights
fastest_flights <- miFlights %>%
  arrange(desc(speed)) %>%
  slice_head(n = 3)
```

3. For flights coming out of GRR, find the 3 flights that traveled the farthest (`distance`) and that arrived the earliest in the morning (`arr_time`) simultaneously.

```{r}
# Filter flights coming out of GRR
flights_from_GRR <- miFlights %>%
  filter(origin == "GRR")

# Find the 3 flights that traveled the farthest (distance) and arrived the earliest in the morning (arr_time)
farthest_and_earliest <- flights_from_GRR %>%
  arrange(desc(distance), arr_time) %>%
  slice_head(n = 3)
```

## The `select()` function

*Drop the year and month columns from miFlights creating a new data set called miDropped.*

```{r}
# Dropping year and month columns
miDropped <- miFlights %>%
  dplyr::select(-year, -month)

miDropped <- miFlights %>%
  dplyr::select(-c(year:month))
```

```{r}
selectExample <- dplyr::select(miFlights, time_hour, air_time, everything(), -day)

selectExample <- miFlights %>%
  dplyr::select(time_hour, air_time, everything(), -day)
```

1. Create a subset of the miFlights data set called timeFlights that only contains variables that end with the word “time”.

```{r}
# Create a subset of miFlights with variables ending with "time"
timeFlights <- miFlights %>%
  select(ends_with("time"))
```

2. Create a new data frame called departureInfo that only has variables that start with “dep”

```{r}
# Create a subset of miFlights with variables starting with "dep"
departureInfo <- miFlights %>%
  select(starts_with("dep"))
```

3. Create a new data frame call newFlights by rearranging the columns of the full miFlights data set so that flight number (flight), origin (origin), and destination (dest) are provided first, then all other columns except the tail number (tailnum).

```{r}
# Rearranging columns in miFlights
newFlights <- miFlights %>%
  dplyr::select(flight, origin, dest, everything(), -tailnum)
```

## The `mutate()` function

```{r}
flights_sml <- miFlights %>% dplyr::select(ends_with("delay"), distance, air_time)

flights_sml <- flights_sml %>% mutate(gain = dep_delay - arr_delay)
```

*Extending the code provided with a single call to `mutate()`, create a new variable, speed, that is equal to distance divided by `air_time`, producing a new data set called flightSpeeds.*

```{r}
flights_sml <- miFlights %>% 
dplyr::select(ends_with("delay"), distance, air_time)

flights_sml <- flights_sml %>% 
  mutate(gain = dep_delay - arr_delay,
         speed = (distance / air_time)*60)
```

*Create a plot showing the relationship between the speed and time gain of each flight, adding appropriate axis and title labels.*

```{r}
# Creating a scatter 
flights_sml %>%
  ggplot(aes(x = speed,
             y = gain,
             color = distance)) +
  geom_point(alpha = 0.1) +
  labs(x = "Speed (miles per hour)",
       y = "Gain (minutes)",
       title = "Michigan flights, 2019 - 2021",
       caption = "Data source: anyFlights R package",
       color = "Distance (miles)") +
  ggthemes::theme_few() +
  theme(legend.position = "bottom",
        legend.text = element_text(angle = 45, hjust = 1))
```

## The `group_by()` and `summarize()` functions

```{r}
## Calculating average departure delay by date
miFlights %>% 
  group_by(year, month, day) %>% 
summarize(delay = mean(dep_delay, na.rm = TRUE)) %>% 
  ungroup() %>%
  slice_head(n = 5)
```

Reproducing the waterfall plot.

```{r}
# Calculating average delay by carrier
delaySummary <- miFlights %>% 
  group_by(carrier_name) %>% 
  summarize(Delay = mean(arr_delay, na.rm = T))
```

```{r}
# Making a waterfall plot (bar plot)
delaySummary %>%
  ggplot(aes(x = fct_reorder(carrier_name, Delay),
             y = Delay,
             fill = Delay > 0)) +
  geom_col() +
  labs(title = "Average flight delay by carrier",
       subtitle = "Michigan flights, 2019 - 2021",
       x = "Carrier" ,
       y = "Average delay (minutes)") +
  scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45,
                                   size = 8, vjust = 0.7))
```

```{r}
# Create a data frame summarizing the median flight delay (arr_delay) by month. Which month has the worst delays? In which month are flights most early / on-time?
monthlyDelays <- miFlights %>% group_by(month) %>%
  summarize(delay = median(arr_delay, na.rm = T))

monthlyDelays %>% knitr::kable()

# Creating a line chart 
monthlyDelays %>% ggplot(aes(x = month,
                             y = delay)) +
  geom_line(color = "dodgerblue") +
  labs(title = "Monthly arrival delays",
       x = "Month",
       y = "Median delay (minutes)" ,
       subtitle = "Michigan flights, 2019-2021",
       caption = "Data source: anyFlights R package") +
  scale_x_continuous(breaks = 1:12) +
  theme_bw()
```

```{r}
# Create a data frame summarizing the median flight delay (arr_delay) by month. Which month has the worst delays? In which month are flights most early / on-time?
monthlyAirportDelays <- miFlights %>% group_by(month, origin) %>%
  summarize(delay = median(arr_delay, na.rm = T))

monthlyAirportDelays %>% knitr::kable()

# Creating a line chart 
monthlyAirportDelays %>% ggplot(aes(x = month,
                      y = delay, color = origin)) +
  geom_line() +
  labs(title = "Monthly arrival delays",
       x = "Month",
       y = "Median delay (minutes)",
       subtitle = "Michigan flights, 2019-2021",
       caption = "Data source: anyFlights R package") +
  scale_x_continuous(breaks = 1:12) +
  theme_bw()
```

# Create a line chart showing the average daily flight delay across time for each of the major airports

```{r}
# Assuming the columns are named year, month, and day
miFlights <- miFlights %>%
  mutate(date = ymd(paste(year, month, day, sep = "-")))

# Now calculating average daily flight delay and create the line chart
averageDailyDelays <- miFlights %>%
  group_by(date, origin) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE))

# Create a line chart
ggplot(averageDailyDelays, aes(x = date, y = avg_delay, color = origin)) +
  geom_line() +
  labs(title = "Average Daily Flight Delay Across Major Airports",
       x = "Date",
       y = "Average Delay (minutes)",
       color = "Airport") +
  theme_minimal()
```

